Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-24 paper.tex

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface (CLI)
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{qiao2012handling, % SpeedGene
deorowicz2013genome, %TGC
sambo2014compression, % snpack
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.
[FIXME need to work SeqArray~\citep{zheng2017seqarray,zheng2012high}
in here.]
The widely-used PLINK binary format stores genotypes in a 
packed binary representation, supporting only biallelic
variants without phase information.
The PLINK 2 PGEN format~\citep{pgen2024} is more general
and compact than PLINK, compressing data using specialised
genotype compression methods~\cite{sambo2014compression}.
Methods have also been developed which store variation data 
along with annotations in databases to facilitate
efficient queries~\cite[e.g.][]{
paila2013gemini,%GEMINI
lopez2017hgva} %OpenCGA
which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
or have storage requirements larger
than uncompressed VCF~\citep{al2023critical}.

>>>>

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression 
and query/compute performance are well 
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods 
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries, 
typically using a command line interface (CLI)
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing 
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{qiao2012handling, % SpeedGene
deorowicz2013genome, %TGC
sambo2014compression, % snpack
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.
[FIXME need to work SeqArray~\citep{zheng2017seqarray,zheng2012high}
in here.]
The widely-used PLINK binary format stores genotypes in a 
packed binary representation, supporting only biallelic
variants without phase information.
The PLINK 2 PGEN format~\citep{pgen2024} is more general
and compact than PLINK, compressing variant data using specialised
algorithms~\cite{sambo2014compression}.
Methods have also been developed which store variation data 
along with annotations in databases to facilitate
efficient queries~\cite[e.g.][]{
paila2013gemini,%GEMINI
lopez2017hgva} %OpenCGA
which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
or have storage requirements larger
than uncompressed VCF~\citep{al2023critical}.
